Beyond body length: Uncovering the drivers of vertebra number variation in Australian blindsnakes

### 2024-11-05 manuscript.Rmd
corrections pre submission

---
# title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
title: "Testing the correlates of vast vertebrae number disparity in Australian blindsnakes"
date: "Version `r Sys.Date()`"
author: ""
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [./custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
csl: evolution.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

@@

We thank museum curators and collection managers from the 
South Australian Museum, Western Australian Museum, Queensland Museum, Australian National Wildlife Collection, Museum and Art Gallery of the Northern Territory, Museums Victoria, and the Florida Museum of Natural History. SH and AA thank Queensland X-ray for provision of imaging for Queensland Museum specimens.
SK thanks the Australian Research Council for ongoing support. ST PhD Scholarship was supported by the Australian Government Research Training Program Scholarship at the Australian National University.

@@

The evolution of the axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates show a lack of pleomerism, suggesting it is possible to dissociate the production of segment and somatic growth, but what factors influence this pattern is not well understood. Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape. Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in the number of vertebrae. We obtained X-ray scans and morphometric data for 42 described species of *Anilios* blindsnakes and evaluated their correlation with each other and ecological variables. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. We also found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate. Our results extend our understanding of the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.

@@

Evolution of the axial skeleton is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [e.g., @taewcharoenSmall2024;@galisFast2014], reptiles [e.g., @vandammeSpeed2002;@molnarMorphological2023], amphibians [e.g., @bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021]. Axial skeleton evolution is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised. Essentially, vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis where the speed of this process relates to the size and number of somites, to produce what will become the axial skeleton [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]. Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975]. While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to the pleomerism rule in moray eels [@reeceEvolutionary2013], large species of snakes [@headDissociation2007;@lindellEvolution1994], salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].

The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveals the developmental processes underlying evolutionary patterns. Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in many snakes [@alexanderPattern1966] and so ventral scales have often been used as a proxy for number of vertebrae [@lindellEvolution1994;@shineVertebral2000]. However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not conform to pleomerism.
Snakes are therefore a good model to study axial evolution because they have great diversity in number of vertebrae and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated the number of vertebrae varies predictably with ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. 
The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. Despite these recent advances, there have been no studies that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for variation and potential ecological drivers.

@@

The Australian blindsnake radiation *Anilios* comprises47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023]. This radiation displays considerable morphological diversity which ranges from narrow-and-long to robust-and-wide species [@tiatragulMorphological2024;@ehmannFamily1993]. However, information about the relationship between body size and number of vertebrae is lacking. In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses about how deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in Australian blindsnakes? (3) Does vertebral ratio vary with ecology (e.g., soil compactness and mean annual temperature)?

@@

We examined 349 specimens representing 48 typhlopid lineages including 44 *Anilios* [@tiatragulAustralia2023;uetzQuarter2021], *Ramphotyphlops multilineatus* (Schlegel, 1839), *R. acuticauda* (Peters, 1878), *R. depressus* (Peters, 1880), and *Acutotyphlops subocularis* (Waite, 1897) (**Table S1**). Our sample size per species ranged from 1 to 25 specimens (median = 6; mean = 7). We sourced specimens from the South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection. Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray radiography. Sex for each specimen was determined by inspecting the gonads or assigned based on relative tail lengths following @tiatragulMorphological2024.

@@

To obtain vertebral counts, we X-ray scanned the specimens on the following machines and institutions: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n  =  87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n = 129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n = 90), and Queensland X-Ray (n = 35).

@@

We selected soil bulk density and annual mean temperature, as we were interested in the potential effects these may have on the distribution of species. Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat a species is most likely living in as they are all strictly fossorial. Mean annual temperature is a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013]. Temperature also directly influences multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018], which influences vertebra development [@martin-del-campoHox2018]. Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restrict penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002]. We extracted soil bulk density values between 0 — 5 cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html) and mean annual temperature from WorldClim v.2 [@fickWorldClim2017] raster files that contained modelled approximations at the resolution of 3 arc-seconds using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022]. We then calculated the species maximum soil bulk density and species mean annual temperature.

@@

All statistical analyses were performed in R Statistical Environment v 4.4.0 [@rcoreteamLanguage2024]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by regressing the species maximum number of total vertebrae against maximum TBL. 
We also calculated the degree of phylogenetic signal from these traits separately using the "physignal.z" function in `geomorph`.
Next, we tested whether the species mean aspect ratio is correlated with mean vertebra ratio (mean number of vertebrae/mean TBL) and mean number of vertebrae. Finally, we tested if variation in vertebra ratio and number of vertebrae is correlated with maximum soil bulk density and mean annual temperature. In the phylogenetic signal and PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. To visualise the relationship between number of vertebrae and intracolumnar vertebral length for each of the representative species, we plotted the mean vertebra size (from three repeated measurements) against the vertebra number. Each vertebra profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].

@@

For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope = 0.28, intercept = 256, R^2^ = 0.02, F = 0.74, Z = 0.30, P = 0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which has the greatest maximum TBL, has relatively few vertebrae. We found phylogenetic signal in both maximum number of total vertebrae (Z = 3.70, P = 0.001, K = 1.2, lambda = 1) and maximum TBL (Z = 3.02, P = 0.002, K = 1.0, lambda = 1).

@@

Our study investigated the presence of pleomerism in Australian blindsnakes and tested whether deviations from pleomerism may be explained by body shape or ecological factors. Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between maximum body length and number of vertebrae reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975]. Variation in body shape, however, is related to vertebra ratio. We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide. We also found that vertebral ratio correlates positively with mean annual temperature and negatively with soil compactness. These results have important implications for understanding the axial evolution of basal snakes and the macroevolutionary dynamics of Australian blindsnakes.

@@

We found a dissociation between the number of vertebrae and maximum adult body length in Australian blindsnake. While pleomerism has been documented in most snake groups [@lindellEvolution1994], exceptionally large species ("giant taxa") within boids, pythonids, and blindsnakes such as *Afrotyphlops schlegelii* have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007]. In line with these observations, some of the largest species in our dataset (*A. torresianus* and *A. pinguis*) possess lower than average number of vertebrae (Fig. \@ref(fig:fig-tbl-vert)). The absence of pleomerism suggests that the large size attained by these "giant taxa" may be due to modification of post-somitogenetic somatic growth [@headDissociation2007]. In other words, species with larger body size do not simply have more vertebra segments, rather the size of each vertebral segment has been elongated during growth. Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004]. Unfortunately, without life-history data for blindsnakes, we cannot test if these mechanisms may be responsible. However, elongation may also be achieved in other ways.
The regionalisation of vertebrae, where distinct regions along the axial column exhibit morphological diversification and evolve relatively independently, may explain the lack of pleomerism in blindsnakes. For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the forebody region compared to larger vertebrae in the hindbody region [@sherrattMacroevolution2022;@sherrattVertebral2019]. Similarly, colubrids have also been shown to have different regionalisation patterns in the vertebral column that contributes to elongation in different ways [@nash-hahnModification2024]. 
The position of the heart is a key marker for delimiting the regions of the snake axial column [@hamptonHeart2022;@hamptonRegionalization2023]. Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.

Lack of pleomerism among burrowing vertebrates is not uncommon. @lindellEvolution1994 found that among Australian elapids and African snakes, burrowing species had fewer vertebrae than expected from their body length compared to terrestrial and arboreal species. The dissociation between number of vertebrae and body size has also been documented in caecilians, a primarily fossorial order of amphibians. In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae, indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*. 
While several studies have documented the number of vertebrae in the highly specialised burrowing amphisbaenianss [@bergmannVertebral2012; @gansAmphisbaeniansreptiles1969; @gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between number of vertebrae and body size in this group. However, they have been shown to deviate from the expected 1:1 ratio of ventral scale number to vertebrae [@alexanderPattern1966], suggesting a lack of pleomerism. We recommend future research to comprehensively investigate the prevalence of pleomerism among burrowing limbless and limb-reduced vertebrates to improve our understanding of the evolution of axial system.

@@

Aspect ratio (body shape) varied with the number of vertebrae in similar ways as previously reported in fish and other snakes (e.g. vipers and colubrids), where species with a more elongated body had relatively more vertebrae than stouter ones [@lindseyPleomerism1975;@lindellEvolution1994;@wardAxial2010]. Two possible explanations have been proposed by @lindellEvolution1994 for snakes including adaptations for locomotion and prey consumption, which are mirrored in the interpretations from elongate fish [@wardAxial2010]. Regarding locomotion, @poughSpecializations1983 showed that in snakes, longer body segments may be advantageous for elongated species that usually move faster, whereas shorter segments are favoured for stouter species that move more slowly. In blindsnakes, however, movement speed may not be as important as how much push force they can generate when burrowing. Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], while species with narrower bodies will be able to use pre-existing small crevices and galleries without needing to expend any energy. In regard to prey consumption, @poughSpecializations1983 suggest that stout species with large heads (i.e. vipers) tend to consume large bulky prey, which could be facilitated by having a stout body as this minimises disruption of physiological functions. However, Australian blindsnakes primarily prey on pupae and larvae of ants and termites, which are relatively small [@webbDietary1993]. As such, we do not expect prey consumption to influence the relationship between aspect ratio and vertebrae we observed. 
<!-- In particular, heavy-bodied blindsnake species (e.g. *A. pinguis*, *A. proximus*, and *A. ligatus*) fed on larger species of ants compared to more slender-bodied blindsnakes species (e.g. *A. hamatus*, *A. bituberculatus*, *A. waitii*, *A. weidii*).  -->

>>>>

---
# title: "Dissociation between number of vertebrae and body size in Australian blindsnakes"
title: "Beyond body length: Uncovering the drivers of vertebra number variation in Australian blindsnakes"
date: "Version `r Sys.Date()`"
author: ""
output: 
  # bookdown::html_document2:
  #   toc: true
  #   number_sections: FALSE
  bookdown::word_document2:
    reference_docx: [./custom-reference-doc.docx]
    number_sections: false
    fig_caption: yes
bibliography: [C:/Users/ST/Documents/repo/anu_thesis_texts/zotero_library.bib]
csl: evolution.csl
link-citations: true
editor_options:
  chunk_output_type: console
---

@@

We thank museum curators and collection managers from the 
South Australian Museum, Western Australian Museum, Queensland Museum, Australian National Wildlife Collection, Museum and Art Gallery of the Northern Territory, Museums Victoria, and the Florida Museum of Natural History. SH and AA thank Queensland X-ray for provision of imaging for Queensland Museum specimens.
JSK thanks the Australian Research Council for ongoing support. ST PhD Scholarship was supported by the Australian Government Research Training Program Scholarship at the Australian National University.

@@

The axial skeleton is important in understanding the evolution of body plans in vertebrates. Changes in the vertebral skeleton can influence body form and modes of locomotion. Elongate animals commonly show a positive correlation between body size and the number of vertebrae, a condition known as pleomerism. However, some groups of vertebrates lack pleomerism, suggesting it is possible to dissociate the production of segmental and somatic growth, but the factors that may influence this pattern are not well understood. Snakes offer a great model to study axial evolution because the number of vertebrae vary greatly between species and they exhibit a wide range of body size and shape. Here, we investigate the prevalence of pleomerism in a continental radiation of fossorial blindsnakes and test whether body shape and ecology can explain the variation in the number of vertebrae. We obtained X-ray scans and morphometric data for 349 museum specimens from 48 lineages of Australo-Papuan blindsnakes and close relatives, and evaluated their correlation with each other and ecological variables. We found an absence of pleomerism among Australian blindsnakes and a correlation between vertebra ratio and body shape, where narrow-and-long species usually possess more vertebrae for their maximum length than wide-and-robust species. We also found that species that have a greater number of vertebrae for their size are generally found in warmer environments and environments with harder-to-penetrate soil substrate. Our results extend our understanding of the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.

@@

Evolution of the axial skeleton is fundamental to vertebrate body plan evolution, facilitating body form changes and influencing modes of locomotion in mammals [e.g., @taewcharoenSmall2024;@galisFast2014], reptiles [e.g., @vandammeSpeed2002;@molnarMorphological2023], amphibians [e.g., @bonettEvidence2017;@adlerEvolutionary2022] and fish [e.g., @yamadaAdaptive2009;@forkerBendy2021]. Axial skeleton evolution is often studied by counting the number of vertebrae because the developmental mechanisms of segmentation are well characterised. Vertebrae begin as somites, epithelial blocks of mesoderm, which form rhythmically during embryogenesis. The speed of this process determines the size and number of somites, producing what will become the axial skeleton  [@gomezControl2008;@dequeantSegmental2008;@gomezDevelopmental2009]. Elongate animals are expected to show a positive correlation between maximum adult body length and number of vertebrae, known as pleomerism [@lindseyPleomerism1975]. While pleomerism was considered the norm [@lindseyPleomerism1975;@wakeComparative1966;@pollyTesting2001], more recent studies have identified exceptions to the pleomerism rule in moray eels [@reeceEvolutionary2013], some clades of snakes [@headDissociation2007;@lindellEvolution1994;@sherrattVertebral2019], and salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].

The lack of pleomerism in some species suggests it is possible to disassociate segment production with body growth, and thus reveal the developmental processes underlying evolutionary patterns. Another clue to disassociation in reptiles can be found in the relationship between the number of ventral scales and precloacal vertebrae – a correlation that is 1:1 in most snakes [@alexanderPattern1966] and so ventral scales are commonly used as a proxy for number of vertebrae [@lindellEvolution1994;@shineVertebral2000]. However, this 1:1 correlation is absent in file snakes, *Achrochodus*, and some blindsnakes, Typhlopidae, [@alexanderPattern1966;@headDissociation2007], as well as some sea snakes of the genus *Hydrophis* [@vorisDermal1975], suggesting these taxa do not show pleomerism.
Snakes are therefore a good model to study axial evolution because they have great diversity in number of vertebrae and a wide range of adult body sizes and shapes but without a limb skeleton to confound signals of locomotor adaptation. Recent studies have indicated evolutionary diversity in number of vertebrae is related to ecological diversification in elapid snakes [@sherrattVertebral2019;@hamptonInterspecific2019;@sherrattMacroevolution2022] and colubrid snakes [@nash-hahnModification2024]. 
The number of vertebrae has also been shown to influence snake locomotion, whereby the number of vertebrae negatively correlates with speed travelled in garter snakes [@kelleyEffects1997]. Despite these recent advances, no studies that have quantified the number of vertebrae across all the species in a diverse assemblage of fossorial snakes to test for variation and potential ecological drivers.

@@

The Australian blindsnake radiation *Anilios* comprises 47 described species distributed across all major biomes of Australia [@tiatragulAustralia2023;@tiatragulPaleoenvironmental2023]. This radiation displays considerable morphological diversity which ranges from narrow-and-long to robust-and-wide species [@tiatragulMorphological2024;@ehmannFamily1993]. However, information about the relationship between body size and number of vertebrae is lacking. In this study, we investigate the prevalence of pleomerism among Australian blindsnakes and test hypotheses about how deviations from pleomerism may be explained by different body shapes or other ecological factors among Australian typhlopid snakes. The specific questions we addressed are: (1) How prevalent is pleomerism among Australian blindsnakes? (2) Can deviations from pleomerism be explained by body shape in Australian blindsnakes? (3) Does vertebral ratio vary with ecology (e.g., soil compactness and mean annual temperature)?

@@

We examined 349 specimens representing 48 typhlopid lineages including 44 *Anilios* [@tiatragulAustralia2023;@uetzQuarter2021], *Ramphotyphlops multilineatus* (Schlegel, 1839), *R. acuticauda* (Peters, 1878), *R. depressus* (Peters, 1880), and *Acutotyphlops subocularis* (Waite, 1897) (**Table S1**). Our sample size per species ranged from 1 to 25 specimens (median = 6; mean = 7). We sourced specimens from the South Australian Museum (SAMA), Western Australian Museum (WAM), Queensland Museum (QM), Museum and Art Gallery of the Northern Territory (MAGNT), and a private collection. Where possible, we chose relatively large non-fragile specimens from both males and females with well-preserved heads for X-ray radiography. Sex for each specimen was determined by inspecting the gonads or assigned based on relative tail lengths following @tiatragulMorphological2024.

@@

To obtain vertebral counts, we X-ray scanned the specimens on the following machines and institutions: LX-60 (Faxitron, IL, USA) at the University of Adelaide Faculty of Health and Medical Sciences (n  =  87), Thermo Kevex PXS5-927EA Microfocus source (Waltham, MA, USA) with a LTX-1717 Digital Flat Panel Detector (Kodex Inc, NJ, USA) at the Western Australian Museum (n = 129), HeliScan MicroCT system at the National Laboratory for X-Ray Micro Computed Tomography (CTLab) based at the Australian National University (n = 90), and FCR0607 (Fujifilm, Tokyo, Japan) at Queensland X-Ray (n = 35).

@@

We selected soil bulk density and annual mean temperature, as we were interested in the potential effects these may have on the distribution of species. Soil bulk density measures the penetrability of soil, which is a proxy for the type of microhabitat a species is most likely living in as they are all strictly fossorial. Mean annual temperature is a climate proxy and is commonly used to describe niche for squamate reptiles [@coxEvolutionary2015;@pieEvolution2017;@wiensDiversity2013]. Temperature also directly influences multiple axes of reptile development [@nobleDevelopmental2018;@boothInfluence2006;@feinerDevelopmental2018], which influences vertebra development [@martin-del-campoHox2018]. Bulk density values > 1.6 g/cm<sup>3</sup> indicate soil that restricts penetration by plant roots, while values <0.5 indicate looser, more organic soil [@mckenzieSoil2002]. We extracted soil bulk density values between 0 — 5 cm depth from [CSIRO Soil and Landscape Grid of Australia](https://www.clw.csiro.au/aclep/soilandlandscapegrid/index.html) and mean annual temperature from WorldClim v.2 [@fickWorldClim2017] raster files that contained modelled approximations at the resolution of 3 arc-seconds using the "extract" function in the R package *`raster`* v.3.6-3 [@hijmansRaster2022]. We then calculated the species maximum soil bulk density and species mean annual temperature.

@@

All statistical analyses were performed in R Statistical Environment v 4.4.0 [@rcoreteamLanguage2024]. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by regressing the species maximum number of total vertebrae against maximum TBL. 
We also calculated the degree of phylogenetic signal from these traits separately using the K metric [@blombergTesting2003] and λ metric [@pagelInferring1999] implemented with the the "physignal.z" function in `geomorph` and "phylosig" function in `phytools` v 2.3 [@revellPhytools2024], respectively. Next, we tested whether the species mean aspect ratio is correlated with mean vertebra ratio (mean number of vertebrae/mean TBL) and mean number of vertebrae. Finally, we tested if variation in vertebra ratio and number of vertebrae is correlated with maximum soil bulk density and mean annual temperature. In the phylogenetic signal and PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. To visualise the relationship between number of vertebrae and intracolumnar vertebral length for each of the representative species, we plotted the mean vertebra size (from three repeated measurements) against the vertebra number. Each vertebra profile was overlaid with a fitted 4° polynomial curve [e.g., @sherrattMacroevolution2022].

@@

For the Australian blindsnake genus *Anilios*, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum TBL (PGLS, slope = 0.28, intercept = 256, R^2^ = 0.02, F = 0.74, Z = 0.30, P = 0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae. Many species have more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) but has the greatest number of vertebrae, while *A. torresianus*, which has the greatest maximum TBL, has relatively few vertebrae. We found more phylogenetic signal than expected under Brownnian motion in both maximum number of total vertebrae (Z = 3.70, P = 0.001, K = 1.2; λ = 1.12, log likelihood(λ) = -197.20, LR(λ=0) = 11.82, P = 0.0006) and maximum TBL (Z = 3.02, P = 0.002, K = 1.0; λ = 1.14, log likelihood(λ) = -221.98, LR(λ=0) = 7.42, P = 0.006).

@@

Our study investigated the presence of pleomerism in Australian blindsnakes and tested whether deviations from pleomerism may be explained by body shape or ecological factors. Our results revealed that pleomerism does not generally apply to Australian typhlopids, indicating a deviation from the positive correlation between maximum body length and number of vertebrae reported for other terrestrial snakes and elongated vertebrates [@headDissociation2007;@lindseyPleomerism1975]. Variation in body shape, however, is related to vertebra ratio. We show that species that are narrow-and-long have proportionally more vertebrae compared to species that are more robust-and-wide. We also found that vertebral ratio correlates positively with mean annual temperature and negatively with soil compactness. These results have important implications for understanding the axial evolution of snakes and the macroevolutionary dynamics of Australian blindsnakes.

@@

We found a dissociation between the number of vertebrae and maximum adult body length in Australian blindsnake. While pleomerism has been documented in most snake groups [@lindellEvolution1994], exceptionally large species ("giant taxa") within boids, pythonids, and blindsnakes such as *Afrotyphlops schlegelii* have fewer than expected pre-cloacal vertebrae for their size [@headDissociation2007]. In line with these observations, some of the largest species in our dataset (*A. torresianus* and *A. pinguis*) possess lower than average number of vertebrae (Fig. \@ref(fig:fig-tbl-vert)). The absence of pleomerism suggests that the large size attained by these "giant taxa" may be due to modification of post-somitogenetic somatic growth [@headDissociation2007]. In other words, species with larger body size do not simply have more vertebra segments, rather the size of each vertebral segment has been elongated during growth. Several mechanisms have been suggested for how giant amniotes achieve their large size including prolonging their growth [@ericksonHow1999] and increasing the rate of growth [@ericksonGigantism2004]. Unfortunately, without life-history data for blindsnakes, we cannot test if these mechanisms may be responsible. However, elongation may also be achieved in other ways.
The regionalisation of vertebrae, where distinct regions along the axial column exhibit morphological diversification and evolve relatively independently, may explain the lack of pleomerism in blindsnakes. For example, the absence of pleomerism in *Hydrophis* sea snakes has been attributed to the increase in the number of smaller vertebrae in the forebody region compared to larger vertebrae in the hindbody region [@sherrattMacroevolution2022;@sherrattVertebral2019]. Similarly, colubrids have also been shown to leverage regionalisation in the vertebral column to achieve elongation in different ways [@nash-hahnModification2024]. 
The position of the heart is a key marker for delimiting the regions of the snake axial column [@hamptonHeart2022;@hamptonRegionalization2023]. Future research should investigate the regionalisation of the vertebral column in blindsnakes as this will provide important insights to understanding the lack of pleomerism.

Lack of pleomerism among burrowing vertebrates is not uncommon. @lindellEvolution1994 found that among Australian elapids and African snakes, burrowing species had fewer vertebrae than expected from their body length compared to terrestrial and arboreal species. The dissociation between number of vertebrae and body size has also been documented in caecilians, a primarily fossorial order of amphibians. In caecilians, @renousBody1989 found that members of the same genus will vary in size but generally share similar number of vertebrae, indicating a lack of pleomerism except for in a few genera such as *Oscaecilia*. 
While several studies have documented the number of vertebrae in the highly specialised burrowing amphisbaenianss [@bergmannVertebral2012; @gansAmphisbaeniansreptiles1969; @gansStudies1960;@gansStudies1971;@gansCharacteristics1978;@alexanderPattern1966], no study has investigated the correlation between number of vertebrae and body size in this group. However, they have been shown to deviate from the expected 1:1 ratio of ventral scale number to vertebrae [@alexanderPattern1966], suggesting a lack of pleomerism. We recommend future research to comprehensively investigate the prevalence of pleomerism among burrowing limbless and limb-reduced vertebrates to improve our understanding of the evolution of the axial system.

@@

Aspect ratio (body shape) varied with the number of vertebrae in similar ways as previously reported in fish and other snakes (e.g. vipers and colubrids), where species with a more elongated body had relatively more vertebrae than stouter ones [@lindseyPleomerism1975;@lindellEvolution1994;@wardAxial2010]. Two possible explanations have been proposed by @lindellEvolution1994 for snakes including adaptations for locomotion and prey consumption, which are mirrored in the interpretations from elongate fish [@wardAxial2010]. Regarding locomotion, @poughSpecializations1983 showed that in snakes, longer body segments may be advantageous for elongated species that usually move faster, whereas shorter segments are favoured for stouter species that move more slowly. In blindsnakes, however, movement speed may not be as important as how much push force they can generate when burrowing. Snakes with relatively wider bodies are generally able to produce greater push force for their body size [@herrelBurrowing2021], while species with narrower bodies will be able to use pre-existing small crevices and galleries with a minimum amount of energy. In regard to prey consumption, @poughSpecializations1983 suggest that stout species with large heads (i.e. vipers) tend to consume large bulky prey, which could be facilitated by having a stout body as this minimises disruption of physiological functions. However, Australian blindsnakes primarily prey on pupae and larvae of ants and termites, which are relatively small [@webbDietary1993]. As such, we do not expect prey consumption to influence the relationship between aspect ratio and vertebrae we observed. 
<!-- In particular, heavy-bodied blindsnake species (e.g. *A. pinguis*, *A. proximus*, and *A. ligatus*) fed on larger species of ants compared to more slender-bodied blindsnakes species (e.g. *A. hamatus*, *A. bituberculatus*, *A. waitii*, *A. weidii*).  -->
